---
title: "p8105_hw2_ls3751"
author: "Liucheng Shi"
output: github_document
---

_packages required_
```{r setup, message = FALSE}
library(tidyverse)
library(readxl)
```


## Problem 1

Read the Mr. Trashwheel dataset. 

```{r, message = F, warning = F}
trashwheel_df = 
	read_xlsx(
		"./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
		sheet = "Mr. Trash Wheel",
		range = cell_cols("A:N")) %>% 
	janitor::clean_names() %>% 
	drop_na(dumpster) %>% 
	mutate(
		sports_balls = round(sports_balls),
		sports_balls = as.integer(sports_balls)
	)
```

Read precipitation data! For 2018 and 2017. 

```{r}
precip_2018 = 
	read_excel(
		"./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
		sheet = "2018 Precipitation",
		skip = 1
	) %>% 
	janitor::clean_names() %>% 
	drop_na(month) %>% 
	mutate(year = 2018) %>% 
	relocate(year)
precip_2017 = 
	read_excel(
		"./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
		sheet = "2017 Precipitation",
		skip = 1
	) %>% 
	janitor::clean_names() %>% 
	drop_na(month) %>% 
	mutate(year = 2017) %>% 
	relocate(year)
```

Now combine annual precipitation dataframes. In the following code chunk, I create a "helper" tibble that contains pairs of numeric and character ways of representing month, and then merge that (using month number as a key) with the precipitation dataset. This technique is one I use often when I need to recode a moderate or large number of values for a variable. 

```{r}
month_df = 
	tibble(
		month = 1:12,
		month_name = month.name
	)
precip_df = 
	bind_rows(precip_2018, precip_2017)
precip_df =
	left_join(precip_df, month_df, by = "month")
```

This dataset contains information from the Mr. Trashwheel trash collector in Baltimore, Maryland. As trash enters the inner harbor, the trashwheel collects that trash, and stores it in a dumpster. The dataset contains information on year, month, and trash collected, include some specific kinds of trash. There are a total of `r nrow(trashwheel_df)` rows in our final dataset. Additional data sheets include month precipitation data. In this dataset:

* The median number of sports balls found in a dumpster in 2017 was `r trashwheel_df %>% filter(year == 2017) %>% pull(sports_balls) %>% median()`
* The total precipitation in 2018 was `r precip_df %>% filter(year == 2018) %>% pull(total) %>% sum()` inches.



## Problem 2

### 2.1 Dataset overview

```{r load the dataset, message = F, warning = F, collapse = T}
nyc_transit_df = read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
                          col_types = "cccddccccccccccccccccclclcccddcc") %>% 
  janitor::clean_names() %>% 
  select(line:entry, vending, ada) %>% 
  mutate(entry = recode(entry, "YES" = T, "NO" = F)) %>% 
  mutate(vending = recode(vending, "YES" = T, "NO" = F))
```

This dataset contains information about NYC subway MTA, including the location, the lines each station served, infrastructure status and so on. The `r ncol(nyc_transit_df)` variables are `r names(nyc_transit_df)`.

The data cleaning process are as follow:

*   read in the dataset and clean the variable names so that they all display as snake case.
*   use select to pull variables of interest, and use mutate (recode) to covert character variables to logical variables.

The dimension of the dataset is `r nrow(nyc_transit_df)` rows * `r ncol(nyc_transit_df)` columns. 

THe dataset is not tidy yet. Routine number and routine name seems to be clustered together that rows did not represent the observations, and columns are not variables.



### 2.2 Distinct stations and ADA compliant

```{r distinguish by name and line, collapse = T}
nyc_distinct = nyc_transit_df %>% 
  distinct(station_name, line, .keep_all = T)

distinct_ada = filter(nyc_distinct, ada == TRUE)

no_vending_entry = filter(nyc_transit_df, vending == F) %>%
                  filter(entry == T)
no_vending = filter(nyc_transit_df, vending == F)
```

*   The number of the __distinct stations__ is `r nrow(nyc_distinct)`.
*   Among them, `r nrow(distinct_ada)` stations are ADA compliant
*   Proportion of stations that have no vending allow entrance is `r count(no_vending_entry)/count(no_vending)`.


### 2.3 Distinct stations serve A & ADA

_Reformat the route number and route name_

```{r}
nyc_transit_tidy = pivot_longer(nyc_distinct, route1:route11,
             names_to = "route_name",
             values_to = "route_number",
             names_prefix = "route")
```

*   There are `r nrow(filter(nyc_transit_tidy, route_number == "A"))` distinct stations serve the A train.  
*   There are `r nrow(filter(nyc_transit_tidy, route_number == "A", ada == TRUE))` distinct stations that A train are ADA compliant.



## Problem 3

### 3.1 Data cleaning Step 1 (pol_month):

*   _break up mon into year, month, and day_.
*   _replace month number with month name, create a president variable_. 
*   _remove prez_dem, prez_gop, and day_.

```{r, message = F}
polmonth_df = read_csv("./data/pols-month.csv") %>% 
  janitor::clean_names() %>% 
  separate(mon, c("year", "month", "day")) %>% 
  mutate(president = case_when(
    prez_gop == 1 ~ "gop",
    prez_dem == 1 ~ "dem"),
         year = as.numeric(year),
         day = as.numeric(day),
         month = as.numeric(month))
polmonth_tidy = polmonth_df %>% 
  left_join(month_df, by = "month") %>% 
  mutate(month = month_name) %>% 
  select(-prez_dem, -prez_gop, -day, -month_name)
```

### 3.2 Data cleaning Step 2 (snp):

*   _use a similar process_.
*   _make year and month leading columns_.

```{r, message = F}
snp_df = read_csv("./data/snp.csv") %>% 
  janitor::clean_names() %>% 
  separate(date, c("month", "day", "year")) %>% 
  mutate(month = as.numeric(month), 
          day = as.numeric(day), 
          year = as.numeric(year))
snp_tidy = snp_df %>% 
  left_join(month_df, by = "month") %>% 
  mutate(month = month_name) %>% 
  select(year, month, close)
```

### 3.3 Data cleaning Step 3 (unemployment):

```{r, message = F}
unemployment_df = read_csv("./data/unemployment.csv") %>% 
  janitor::clean_names() %>% 
  pivot_longer(jan:dec,
    names_to = "month",
    values_to = "unemployment") %>% 
  mutate(year = as.integer(year))
```

```{r}
month_Lower = tibble(
  month = c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec"),
  month_name = month.name
)
```

```{r}
unemployment_tidy = unemployment_df %>% 
  left_join(month_Lower, by = "month") %>% 
  mutate(month = month_name) %>% 
  select(year, month, unemployment)
```

_Join the three datasets accordingly_

```{r}
analy_df = 
  left_join(polmonth_tidy,snp_tidy, by = c("year","month")) %>% 
  left_join(unemployment_tidy, by = c("year","month"))
```
